---
output: rmarkdown::github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# modernGLMM

[![R >= 4.1.0](https://img.shields.io/badge/R-%3E%3D%204.1.0-276DC3.svg)](https://www.r-project.org/)
[![License: GPL-3](https://img.shields.io/badge/License-GPL--3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0.en.html)
[![Lifecycle: active](https://img.shields.io/badge/lifecycle-active-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#active)

`modernGLMM` provides unified, reproducible R implementations of examples from:

> Stroup, W. W., Ptukhina, M., and Garai, S. (2024). *Generalized Linear Mixed Models: Modern Concepts, Methods and Applications* (2nd ed.). CRC Press.

The book uses SAS for its worked examples and describes why a single R implementation was not supplied: the R GLMM ecosystem is broad, capable, and fragmented across specialized packages. `modernGLMM` addresses that practical gap by collecting coherent R workflows for the examples and by documenting how the pieces fit together.

The package is intended for teaching, applied statistical modeling, and reproducible verification. It keeps the book's model structure as the reference point while using modern R tools for estimation, diagnostics, post hoc inference, reporting, and visualization.

## Scope

`modernGLMM` includes datasets, example scripts, vignettes, and verification artifacts for workflows spanning:

- Gaussian linear and mixed models
- Generalized linear models and generalized linear mixed models
- BLUPs and predictable functions
- Treatment structures and multi-level designs
- Counts, rates, proportions, multinomial responses, and zero-inflated models
- Time-to-event data, smoothing splines, repeated measures, correlated errors, Bayesian GLMMs, and power analysis

The R implementation layer uses `lme4`, `glmmTMB`, `nlme`, `mgcv`, `gamm4`, `brms`, `survival`, `coxme`, `emmeans`, `DHARMa`, `report`, and related packages where each tool is appropriate.

## Installation

Install the development version from GitHub:

```{r, eval = FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}
remotes::install_github("myaseen208/modernGLMM")
```

After CRAN release, install with:

```{r, eval = FALSE}
install.packages("modernGLMM")
```

## Example

```{r, eval = FALSE}
library(modernGLMM)

data(DataSet11.1)

fit <- stats::glm(
  y ~ trt,
  family = stats::poisson(link = "log"),
  data = DataSet11.1
)

summary(fit)

if (requireNamespace("emmeans", quietly = TRUE)) {
  emmeans::emmeans(fit, specs = ~ trt, type = "response")
}

if (requireNamespace("report", quietly = TRUE)) {
  report::report(fit)
}
```

## Verification Status

Verification artifacts are stored in `inst/verification/`. Current verification covers Chapter 1, Chapters 2-5, Chapters 8-10, Chapter 11, and Chapter 12-21 legacy/no-export classifications. Full numerical reproduction is claimed only for rows marked `EXACT` or `APPROX`.

## Citation

```{r, eval = FALSE}
citation("modernGLMM")
```
