---
title: "Chapter 12: Rates and Proportions"
subtitle: "Beta Regression and Binomial GLMMs for Rates"
author: "Muhammad Yaseen"
date: last-modified
vignette: >
  %\VignetteIndexEntry{Chapter 12: Rates and Proportions}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: show
    df-print: kable
    fig-width: 7
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(modernGLMM)
library(lme4)
library(emmeans)
library(ggplot2)
```

## Overview

Chapter 12 addresses responses that are **proportions or rates** — either
continuous values strictly in $(0, 1)$ (Section 12.6) or binomial counts
with a fixed denominator (Section 12.3).

Two primary approaches:

1. **Beta GLMM**: models a continuous proportion using the Beta distribution
   with a logit link; accommodates between-run variability via random effects
2. **Binomial GLMM**: models the number of successes out of a fixed total
   with a nested or split-plot random-effect structure

## Example 12.1 — Continuous Proportions: Beta GLMM (Section 12.6.2)

`DataSet12.1`: 2 treatments × 12 runs × 6 dose levels = 144 observations.
Response is a continuous proportion in $(0, 1)$.

```{r}
data(DataSet12.1)
str(DataSet12.1)
## Mean proportion by treatment and dose
with(DataSet12.1, tapply(proportion, list(trt, dose), mean))
```

```{r}
#| label: fig-12-1-dose
#| fig-cap: "Dose-response profiles by treatment"
ggplot(DataSet12.1, aes(x = dose, y = proportion,
                         colour = trt, group = trt)) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  labs(title = "Chapter 12: Dose-response by treatment",
       x = "Dose level", y = "Mean proportion",
       colour = "Treatment") +
  theme_bw()
```

### Beta GLMM via glmmTMB

```{r}
if (requireNamespace("glmmTMB", quietly = TRUE)) {
  fit_beta <- glmmTMB::glmmTMB(
    proportion ~ trt * dose + (1 | run),
    family  = glmmTMB::beta_family(link = "logit"),
    data    = DataSet12.1
  )
  summary(fit_beta)
}
```

```{r}
if (requireNamespace("glmmTMB", quietly = TRUE)) {
  emm_beta <- emmeans::emmeans(fit_beta, ~ trt | dose,
                                at = list(dose = c(0, 5)),
                                type = "response")
  print(emm_beta)
  emmeans::contrast(emm_beta, method = "pairwise")
}
```

## Example 12.2 — Binomial Nested Factorial (Section 12.3.2)

`DataSet12.2`: 10 blocks × 6 treatments (factor A with 2 levels, factor B
with 3 levels nested within A) = 60 observations.
Response is count of successes `f` out of total `n`.

```{r}
data(DataSet12.2)
str(DataSet12.2)
## Observed proportions by A and B
with(DataSet12.2, tapply(f / n, list(a, b), mean))
```

```{r}
fit_nb <- lme4::glmer(
  cbind(f, n - f) ~ a / b + (1 + a | block),
  family  = stats::binomial(link = "logit"),
  data    = DataSet12.2,
  control = lme4::glmerControl(optimizer = "bobyqa")
)
summary(fit_nb)
```

```{r}
if (requireNamespace("car", quietly = TRUE)) {
  car::Anova(fit_nb, type = "III")
}
```

```{r}
## B(A) means on probability scale
emm_ba <- emmeans::emmeans(fit_nb, ~ b | a, type = "response")
print(emm_ba)
```

## Comparing Approaches

| Approach | Distribution | When to use |
|---|---|---|
| Beta GLMM | Beta$(μφ, (1-μ)φ)$ | Continuous proportions in $(0,1)$; run-level random effects |
| Binomial GLMM | Binomial$(n, π)$ | Count/total with known $n$; factorial or nested designs |

## Key Takeaways

- **Beta GLMM** is the appropriate model for continuous proportions.
- **Binomial GLMM** is correct when the denominator $n$ is a fixed count.
- Both models use the logit link, making linear predictors directly
  comparable.
- Random effects (run, block) account for over-dispersion and design structure.

## References

Stroup, W. W., Ptukhina, M., and Garai, S. (2024). *Generalized Linear Mixed Models: Modern Concepts, Methods and Applications* (2nd ed.). CRC Press.

Ferrari, S., & Cribari-Neto, F. (2004). Beta regression for modelling
rates and proportions. *Journal of Applied Statistics*, 31(7), 799–815.
