---
title: "Chapter 7: Inference, Part II"
subtitle: "Fixed-Effects Inference in Linear Mixed Models"
author: "Muhammad Yaseen"
date: today
vignette: >
  %\VignetteIndexEntry{Chapter 7: Inference, Part II}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    df-print: kable
    fig-width: 7
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(modernGLMM)
library(emmeans)
library(car)
```

## Overview

Chapter 7 covers fixed-effects inference: constructing estimable functions,
computing contrasts, and performing multiple comparisons in (G)LMMs.

Key topics:

- **Estimability**: not all linear combinations of $\boldsymbol{\beta}$ can be
  estimated; `emmeans` handles this transparently.
- **Contrasts**: custom linear combinations via `emmeans::contrast()`.
- **Multiple comparisons**: Tukey, Bonferroni, Scheffé, and FDR adjustments.
- **Interaction dissection**: `emmeans::joint_tests()` for simple effects.

## Treatment Structure Example (DataSet8.1)

The 2×3 factorial from Chapter 8 is used here to illustrate contrast coding.

```{r}
data(DataSet8.1)
DataSet8.1$a <- factor(DataSet8.1$a)
DataSet8.1$b <- factor(DataSet8.1$b)

fit <- stats::lm(y ~ a * b, data = DataSet8.1)
```

### Estimated marginal means

```{r}
emm <- emmeans::emmeans(fit, ~ a * b)
print(emm)
```

### All pairwise comparisons within each level of `a`

```{r}
pairs(
  emmeans::emmeans(fit, ~ b | a),
  adjust = "tukey"
)
```

### Simple effects (ANOVA for b within each level of a)

```{r}
emmeans::joint_tests(fit, by = "a")
```

### Custom contrasts

```{r}
emmeans::contrast(emm, list(
  "a0: b0 vs b1" = c(1, -1, 0, 0, 0, 0),
  "a1: b0 vs b1" = c(0, 0, 0, 1, -1, 0)
))
```

## Key Takeaways

- `emmeans::emmeans()` computes estimable marginal means automatically.
- `emmeans::contrast()` accepts named lists of contrast vectors.
- `emmeans::joint_tests()` provides F-tests for each term, optionally `by`
  a conditioning variable (equivalent to SAS `SLICE`).
- Multiplicity adjustment (`adjust = "tukey"`, `"bonferroni"`, `"fdr"`) is
  passed as an argument to `pairs()` or `contrast()`.

## References

Stroup, W. W., Ptukhina, M., and Garai, S. (2024). *Generalized Linear Mixed
Models: Modern Concepts, Methods and Applications (2nd ed.)*. CRC Press.
