---
title: "Chapter 4: Pre-GLMM Estimation and Inference Basics"
subtitle: "OLS, GLS, BLUE, Estimable Functions, and Quadratic Forms"
author: "Muhammad Yaseen"
date: last-modified
vignette: >
  %\VignetteIndexEntry{Chapter 4: Pre-GLMM Estimation and Inference Basics}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    df-print: kable
    fig-width: 7
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

## Overview

Chapter 4 of Stroup, Ptukhina & Garai (2024) establishes the
**pre-GLMM estimation and inference framework** that underlies
the mixed model machinery developed in subsequent chapters.
This chapter is entirely theoretical — it contains no applied examples
or datasets.  The core topics are:

- Ordinary Least Squares (OLS) and Generalised Least Squares (GLS)
- Best Linear Unbiased Estimators (BLUE)
- Estimable functions and testable hypotheses
- Quadratic forms and their distributions

## Ordinary Least Squares

For a linear model $\mathbf{y} = \mathbf{X}\pmb{\beta} + \pmb{\varepsilon}$
with $\pmb{\varepsilon} \sim \mathcal{N}(\mathbf{0}, \sigma^2 \mathbf{I})$,
the OLS estimator is:

$$\hat{\pmb{\beta}}_{\text{OLS}} = (\mathbf{X}^\top \mathbf{X})^{-} \mathbf{X}^\top \mathbf{y}$$

where $(-)$ denotes a generalised inverse.  OLS is BLUE when errors are
homoscedastic and uncorrelated (Gauss-Markov theorem).

## Generalised Least Squares

When $\text{Var}(\pmb{\varepsilon}) = \sigma^2 \mathbf{V}$ with
$\mathbf{V} \ne \mathbf{I}$, GLS applies:

$$\hat{\pmb{\beta}}_{\text{GLS}} = (\mathbf{X}^\top \mathbf{V}^{-1} \mathbf{X})^{-}
\mathbf{X}^\top \mathbf{V}^{-1} \mathbf{y}$$

GLS is BLUE under the correctly specified $\mathbf{V}$.  In practice
$\mathbf{V}$ is unknown and must be estimated, leading to the REML/ML
estimation machinery of Chapter 5.

## Best Linear Unbiased Estimators (BLUE)

An estimator $\hat{\pmb{\ell}} = \mathbf{c}^\top \hat{\pmb{\beta}}$
is BLUE for $\pmb{\ell} = \mathbf{c}^\top \pmb{\beta}$ if:

1. It is **linear** in $\mathbf{y}$.
2. It is **unbiased**: $\mathsf{E}[\hat{\pmb{\ell}}] = \pmb{\ell}$.
3. It has **minimum variance** among all linear unbiased estimators.

The GLS estimator achieves this under the correct $\mathbf{V}$.

## Estimable Functions

A linear combination $\mathbf{c}^\top \pmb{\beta}$ is **estimable**
if and only if $\mathbf{c}^\top$ lies in the row space of $\mathbf{X}$,
i.e., $\mathbf{c}^\top = \mathbf{t}^\top \mathbf{X}$ for some $\mathbf{t}$.

Non-estimable combinations — arising from rank deficiency or aliased
parameters — cannot be uniquely estimated regardless of the data, and
any hypothesis test involving them is not meaningful.  The `emmeans`
package performs estimability checks automatically.

```{r}
## Illustration: rank-deficient design leads to non-estimable contrasts
X <- matrix(c(1,1,0, 1,0,1, 1,1,0), nrow = 3, byrow = TRUE)
rankX <- qr(X)$rank
cat("Rank of X:", rankX, "  (columns:", ncol(X), ")\n")
cat("Columns span a", rankX, "-dimensional space —",
    ncol(X) - rankX, "parameter(s) are not independently estimable.\n")
```

## Quadratic Forms

Hypothesis tests in linear models rely on **quadratic forms**
$\mathbf{y}^\top \mathbf{A} \mathbf{y}$.  Key results:

- If $\mathbf{y} \sim \mathcal{N}(\pmb{\mu}, \sigma^2 \mathbf{I})$
  and $\mathbf{A}$ is idempotent of rank $r$, then
  $\mathbf{y}^\top \mathbf{A} \mathbf{y} / \sigma^2 \sim \chi^2_r(\lambda)$
  with non-centrality $\lambda = \pmb{\mu}^\top \mathbf{A} \pmb{\mu} / \sigma^2$.
- The F-statistic is a ratio of two chi-squared quadratic forms
  divided by their respective degrees of freedom.

```{r}
## Non-centrality parameter for a one-way ANOVA (k=3, n=5, sigma^2=9)
mu    <- c(20, 25, 25)
mubar <- mean(mu)
sigma2 <- 9
n      <- 5
lambda <- n * sum((mu - mubar)^2) / sigma2
cat("Non-centrality parameter lambda =", round(lambda, 5), "\n")

## Power at F_{2,12} critical value
df1 <- 2; df2 <- 12
Fc  <- stats::qf(0.95, df1, df2)
pow <- 1 - stats::pf(Fc, df1, df2, ncp = lambda)
cat("Power (n = 5):", round(pow, 5), "\n")
```

## Key Takeaways

- OLS is BLUE only when errors are i.i.d.; correlated or heteroscedastic
  errors require GLS.
- Estimability is a prerequisite for meaningful inference — `emmeans`
  checks this automatically.
- Quadratic form theory underpins all ANOVA and F-tests in the LMM/GLMM
  framework developed in Chapters 6–21.
- Chapter 4 has no applied examples; its applied manifestations appear
  in Chapters 8–10 (Gaussian LMMs) and beyond.

## References

Stroup, W. W., Ptukhina, M., and Garai, S. (2024). *Generalized Linear Mixed
Models: Modern Concepts, Methods and Applications* (2nd ed.). CRC Press.
Chapter 4: Pre-GLMM Estimation and Inference Basics, pp. 137–145.
