---
title: "Chapter 9: Multi-Level Models"
subtitle: "Nested Factorial, Split-Plot, and Blocked Designs"
author: "Muhammad Yaseen"
date: last-modified
vignette: >
  %\VignetteIndexEntry{Chapter 9: Multi-Level Models}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: show
    df-print: kable
    fig-width: 7
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(modernGLMM)
library(lme4)
library(lmerTest)
library(emmeans)
```

## Overview

Chapter 9 extends the GLMM framework to **binary and binomial responses**.
A binary observation $y_i \in \{0, 1\}$ follows a Bernoulli distribution;
a binomial aggregate $y_i \sim \text{Binomial}(n_i, p_i)$ counts successes
among $n_i$ trials.

The canonical link for both is the **logit**:

$$\eta_i = \text{logit}(p_i) = \log\frac{p_i}{1-p_i} = \mathbf{x}_i^\top\pmb{\beta} + \mathbf{z}_i^\top\mathbf{b}$$

with $\mathbf{b} \sim \mathcal{N}(\mathbf{0}, \mathbf{G})$.

## Example 9.1 — Nested Factorial Structure

Dataset: 3 sets, 2 treatments per set, arranged in blocks nested within sets.

```{r}
data(DataSet9.1)
DataSet9.1$block <- factor(DataSet9.1$block)
DataSet9.1$set   <- factor(DataSet9.1$set)
DataSet9.1$trt   <- factor(DataSet9.1$trt)
str(DataSet9.1)
```

### Fit the LMM (Gaussian approximation)

```{r}
Exam9.1Lmer <- lmerTest::lmer(
  y ~ set + set:trt + (1 | set:block),
  data    = DataSet9.1,
  control = lme4::lmerControl(optimizer = "bobyqa")
)
summary(Exam9.1Lmer)
stats::anova(Exam9.1Lmer, ddf = "Kenward-Roger")
```

### Estimated Marginal Means

```{r}
emm9.1 <- emmeans::emmeans(Exam9.1Lmer, ~ trt | set)
print(emm9.1)

emmeans::contrast(emm9.1, method = "pairwise", by = "set")
```

### Interpretation

```{r}
if (requireNamespace("report", quietly = TRUE)) {
  report::report(Exam9.1Lmer)
}
```

## Example 9.2 — Split-Plot with Row-Column Structure

```{r}
data(DataSet9.2)
DataSet9.2$block <- factor(DataSet9.2$block)
DataSet9.2$row   <- factor(DataSet9.2$row)
DataSet9.2$col   <- factor(DataSet9.2$col)
DataSet9.2$a     <- factor(DataSet9.2$a)
DataSet9.2$b     <- factor(DataSet9.2$b)
str(DataSet9.2)
```

```{r}
Exam9.2Lmer <- lmerTest::lmer(
  y ~ a * b + (1 | block) + (1 | block:a) + (1 | block:b),
  data    = DataSet9.2,
  control = lme4::lmerControl(optimizer = "bobyqa")
)
summary(Exam9.2Lmer)
stats::anova(Exam9.2Lmer, ddf = "Kenward-Roger")
```

```{r}
emm9.2 <- emmeans::emmeans(Exam9.2Lmer, ~ a * b)
print(emm9.2)
emmeans::contrast(emm9.2, method = "pairwise",
                  by = "a", adjust = "tukey")
```

## Example 9.3 — Factorial with Blocks

```{r}
data(DataSet9.3)
DataSet9.3$block <- factor(DataSet9.3$block)
DataSet9.3$aa    <- factor(DataSet9.3$a)
DataSet9.3$bb    <- factor(DataSet9.3$b)
DataSet9.3$cc    <- factor(DataSet9.3$c)
DataSet9.3$a2    <- DataSet9.3$a^2
DataSet9.3$b2    <- DataSet9.3$b^2
DataSet9.3$c2    <- DataSet9.3$c^2
DataSet9.3$ab    <- DataSet9.3$a * DataSet9.3$b
DataSet9.3$ac    <- DataSet9.3$a * DataSet9.3$c
DataSet9.3$bc    <- DataSet9.3$b * DataSet9.3$c
str(DataSet9.3)
```

```{r}
Exam9.3Lmer <- lmerTest::lmer(
  y ~ a + a2 + b + b2 + c + bc + (1 | block),
  data    = DataSet9.3,
  control = lme4::lmerControl(optimizer = "bobyqa")
)
summary(Exam9.3Lmer)
stats::anova(Exam9.3Lmer, ddf = "Kenward-Roger", type = 1)
```

## Overdispersion Diagnostics

```{r}
if (requireNamespace("performance", quietly = TRUE)) {
  tryCatch(
    performance::check_model(Exam9.1Lmer),
    error = function(e) message("check_model() error: ", conditionMessage(e))
  )
}
```

```{r}
if (requireNamespace("DHARMa", quietly = TRUE)) {
  sim_r <- DHARMa::simulateResiduals(Exam9.1Lmer, plot = TRUE)
}
```

## Key Takeaways

- **Nested and split-plot structures** require matching random effects
  specifications to the design structure.
- The `(1 | set:block)` term captures blocks nested within sets without adding an unprinted random set variance component.
- `emmeans` provides estimated marginal means with proper standard errors
  that account for the random effects structure.
- DHARMa diagnostics help identify departures from model assumptions.

## References

Stroup, W. W., Ptukhina, M., and Garai, S. (2024). *Generalized Linear Mixed Models: Modern Concepts, Methods and Applications* (2nd ed.). CRC Press.
