---
title: "Chapter 11: Counts"
author: "Muhammad Yaseen"
date: today
vignette: >
  %\VignetteIndexEntry{Chapter 11: Counts}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    number-sections: true
    theme: cosmo
    code-fold: show
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Count Models

Chapter 11 develops GLMMs for non-negative integer responses. The Poisson
model assumes

$$
Y_i \mid b_i \sim \operatorname{Poisson}(\lambda_i),
\qquad
\operatorname{E}(Y_i \mid b_i) = \operatorname{Var}(Y_i \mid b_i) = \lambda_i.
$$

With a log link,

$$
\log(\lambda_i) = x_i^\top \beta + z_i^\top b.
$$

Overdispersion is present when the conditional variance is larger than the
Poisson variance. A negative-binomial GLMM adds a scale parameter:

$$
Y_i \mid b_i \sim \operatorname{NB}(\lambda_i, \phi),
\qquad
\operatorname{Var}(Y_i \mid b_i) = \lambda_i(1 + \phi\lambda_i).
$$

## Example 11.1

Section 11.1.2 compares pre-GLM ANOVA methods and a Poisson-normal GLMM for a
two-treatment completely randomized count dataset.

```{r}
data(DataSet11.1, package = "modernGLMM")
utils::str(DataSet11.1)
stats::aggregate(count ~ trt, data = DataSet11.1, FUN = mean)
```

### ANOVA on Counts

```{r}
fit_count <- stats::lm(count ~ trt, data = DataSet11.1)
stats::anova(fit_count)
emmeans::emmeans(fit_count, specs = ~ trt)
```

The reconstructed counts reproduce the printed treatment means, residual mean
square, F statistic, and p-value for the untransformed ANOVA.

### ANOVA on Log Counts

```{r}
DataSet11.1$log_count <- log(DataSet11.1$count)
fit_log <- stats::lm(log_count ~ trt, data = DataSet11.1)
stats::anova(fit_log)
emmeans::emmeans(fit_log, specs = ~ trt)
```

### Poisson GLM

```{r}
fit_pois_glm <- stats::glm(
  count ~ trt,
  family = stats::poisson(link = "log"),
  data = DataSet11.1
)

emmeans::emmeans(fit_pois_glm, specs = ~ trt, type = "response")
sum(stats::residuals(fit_pois_glm, type = "pearson")^2) /
  stats::df.residual(fit_pois_glm)
```

The Pearson chi-square/DF is much larger than one, indicating that the Poisson
GLM has not accounted for the unit-level source of variation.

### Poisson-Normal GLMM

```{r}
fit_pois_normal <- lme4::glmer(
  count ~ trt + (1 | trt:unit),
  family = stats::poisson(link = "log"),
  data = DataSet11.1,
  nAGQ = 0,
  control = lme4::glmerControl(optimizer = "bobyqa")
)

summary(fit_pois_normal)
lme4::VarCorr(fit_pois_normal)
emmeans::emmeans(fit_pois_normal, specs = ~ trt, type = "response")
```

```{r}
if (requireNamespace("DHARMa", quietly = TRUE)) {
  sim_11_1 <- DHARMa::simulateResiduals(fit_pois_normal, plot = FALSE)
  DHARMa::testDispersion(sim_11_1)
}

if (requireNamespace("report", quietly = TRUE)) {
  report::report(fit_pois_normal)
}
```

## Example 11.3

Section 11.4 uses a randomized complete block design with three treatments and
ten blocks to show how overdispersion changes inference.

```{r}
data(DataSet11.3, package = "modernGLMM")
utils::str(DataSet11.3)
stats::aggregate(count ~ trt, data = DataSet11.3, FUN = mean)
```

### Naive Poisson GLMM

```{r}
fit_naive <- lme4::glmer(
  count ~ trt + (1 | block),
  family = stats::poisson(link = "log"),
  data = DataSet11.3,
  nAGQ = 1,
  control = lme4::glmerControl(optimizer = "bobyqa")
)

pearson_df <- sum(stats::residuals(fit_naive, type = "pearson")^2) /
  stats::df.residual(fit_naive)
pearson_df
emmeans::emmeans(fit_naive, specs = ~ trt, type = "response")
```

The diagnostic remains high, so the naive model should not be used for final
inference.

### Poisson-Normal GLMM

```{r}
fit_unit <- lme4::glmer(
  count ~ trt + (1 | block) + (1 | block:trt),
  family = stats::poisson(link = "log"),
  data = DataSet11.3,
  nAGQ = 0,
  control = lme4::glmerControl(optimizer = "bobyqa")
)

lme4::VarCorr(fit_unit)
emmeans::emmeans(fit_unit, specs = ~ trt, type = "response")
```

The block-by-treatment random term is the conditional-model analogue of the
unit-level source of variation in the repurposed ANOVA.

### Negative-Binomial GLMM

```{r}
if (requireNamespace("glmmTMB", quietly = TRUE)) {
  fit_nb <- glmmTMB::glmmTMB(
    count ~ trt + (1 | block),
    family = glmmTMB::nbinom2(link = "log"),
    data = DataSet11.3
  )

  summary(fit_nb)
  emmeans::emmeans(fit_nb, specs = ~ trt, type = "response")
  1 / sigma(fit_nb)
}
```

```{r}
if (requireNamespace("glmmTMB", quietly = TRUE) &&
    requireNamespace("DHARMa", quietly = TRUE)) {
  sim_11_3 <- DHARMa::simulateResiduals(fit_nb, plot = FALSE)
  DHARMa::testDispersion(sim_11_3)
}

if (requireNamespace("glmmTMB", quietly = TRUE) &&
    requireNamespace("report", quietly = TRUE)) {
  report::report(fit_nb)
}
```

## Example 11.4

Section 11.5 considers a multi-level split-plot design with a 7 by 4 factorial
treatment structure. The package data are synthetic because the original
`sp_counts` file was not available.

```{r}
data(DataSet11.4, package = "modernGLMM")
utils::str(DataSet11.4)
stats::aggregate(count ~ a + b, data = DataSet11.4, FUN = mean)
```

```{r}
if (requireNamespace("glmmTMB", quietly = TRUE)) {
  fit_sp_nb <- glmmTMB::glmmTMB(
    count ~ a * b + (1 | block) + (1 | block:a),
    family = glmmTMB::nbinom2(link = "log"),
    data = DataSet11.4
  )

  summary(fit_sp_nb)
  emm_sp <- emmeans::emmeans(fit_sp_nb, specs = ~ a * b, type = "response")
  emm_sp
}
```

```{r}
if (requireNamespace("glmmTMB", quietly = TRUE)) {
  emm_df <- as.data.frame(emm_sp)
  ggplot2::ggplot(
    emm_df,
    ggplot2::aes(x = b, y = response, group = a, colour = a)
  ) +
    ggplot2::geom_line(linewidth = 0.8) +
    ggplot2::geom_point(size = 2) +
    ggplot2::labs(x = "Factor B", y = "Estimated count", colour = "Factor A") +
    ggplot2::theme_bw()
}
```

## Chapter Summary

Poisson GLMMs are useful starting points for count data, but their
mean-variance restriction is often too rigid. The practical workflow is to fit
a model that matches the design skeleton, check overdispersion, and then choose
an inference target. Conditional inference uses random effects and
distributions such as Poisson-normal or negative binomial; marginal inference
uses working covariance or quasi-likelihood machinery.

