---
title: "Chapter 19: Bayesian Implementation of GLMM"
subtitle: "Bayesian Estimation with brms and Stan"
author: "Muhammad Yaseen"
date: today
vignette: >
  %\VignetteIndexEntry{Chapter 19: Bayesian Implementation of GLMM}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: [cosmo, theme.scss]
    df-print: kable
    fig-width: 7
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(modernGLMM)
library(lme4)
library(emmeans)
```

## Overview

Chapter 19 introduces the **Bayesian approach** to GLMMs. The Bayesian
framework treats all unknowns

- fixed effects $\boldsymbol{\beta}$,random effects $\mathbf{b}$, and variance components $\boldsymbol{\theta}$
- as random variables with prior distributions.

**Bayes' Theorem** for mixed models:

$$p(\boldsymbol{\beta}, \mathbf{b}, \boldsymbol{\theta} \mid \mathbf{y}) \propto p(\mathbf{y} \mid \boldsymbol{\beta}, \mathbf{b}, \boldsymbol{\theta}) \cdot p(\boldsymbol{\beta}) \cdot p(\mathbf{b} \mid \boldsymbol{\theta}) \cdot p(\boldsymbol{\theta})$$

Key advantages over REML/ML:

- Full posterior distributions for all parameters (not just point estimates)
- Exact small-sample inference (no asymptotic approximations)
- Natural propagation of uncertainty to predictions
- Handles complex variance structures and non-standard models

## Prior Specification

Typical weakly informative priors for GLMMs:

```{r}
#| eval: false
# Using brms (requires Stan backend)
if (requireNamespace("brms", quietly = TRUE)) {
  priors <- brms::prior(normal(0, 10), class = b) +
            brms::prior(normal(0, 10), class = Intercept) +
            brms::prior(cauchy(0, 2.5), class = sd)
}
```

## Bayesian One-Way Random Effects Model

```{r}
if (requireNamespace("brms", quietly = TRUE)) {
  data(DataSet10.1)
  DataSet10.1$a <- factor(DataSet10.1$a)

  # Frequentist reference
  fit_freq <- lme4::lmer(y ~ 1 + (1 | a), data = DataSet10.1)
  cat("REML estimates:\n")
  print(as.data.frame(lme4::VarCorr(fit_freq)))

  # Bayesian equivalent
  fit_bayes <- brms::brm(
    y ~ 1 + (1 | a),
    data   = DataSet10.1,
    prior  = brms::prior(normal(0, 10), class = Intercept) +
             brms::prior(cauchy(0, 2.5), class = sd),
    chains = 4,
    iter   = 2000,
    seed   = 42,
    silent = 2
  )
  summary(fit_bayes)
}
```

## Posterior Predictive Check

```{r}
if (requireNamespace("brms", quietly = TRUE)) {
  brms::pp_check(fit_bayes, ndraws = 50)
}
```

## Posterior for Variance Components

```{r}
if (requireNamespace("brms", quietly = TRUE)) {
  brms::mcmc_plot(fit_bayes, variable = c("sd_a__Intercept", "sigma"),
                  type = "hist")
}
```

## Key Takeaways

- `brms::brm()` fits Bayesian GLMMs with the same formula syntax as `lme4`.
- Posterior distributions replace point estimates + standard errors.
- `brms::pp_check()` provides graphical posterior predictive checks.
- Weakly informative priors (`normal(0, 10)`, `cauchy(0, 2.5)`) stabilise
  estimation without strongly influencing the posterior when data are informative.

## References

Stroup, W. W., Ptukhina, M., and Garai, S. (2024). *Generalized Linear Mixed
Models: Modern Concepts, Methods and Applications (2nd ed.)*. CRC Press.
