---
title: "Chapter 17: Correlated Errors, Part I: Repeated Measures"
subtitle: "Covariance Structures for Correlated Observations"
author: "Muhammad Yaseen"
date: last-modified
vignette: >
  %\VignetteIndexEntry{Chapter 17: Correlated Errors, Part I: Repeated Measures}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: show
    df-print: kable
    fig-width: 8
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(modernGLMM)
library(nlme)
library(emmeans)
library(ggplot2)
```

## Overview

Chapter 17 addresses **repeated measures designs** where the same subject
is observed multiple times.  Key challenges:

- Observations on the same subject are **correlated**
- The correlation structure needs to be specified or estimated
- Standard ANOVA assuming independence is incorrect

### Common Covariance Structures

| Name | Symbol | Description |
|---|---|---|
| Compound Symmetric | CS | Equal correlation $\rho$ between all pairs |
| AR(1) | AR1 | Correlation $\rho^{\|t-t'\|}$ decays with lag |
| ARH(1) | ARH1 | AR(1) with heterogeneous variances per time |
| SP(POW) | SPPOW | Power model for unequally-spaced times |

## Example 17.1 — Crossover with ARH(1) (Section 17.3.1)

`DataSet17.1`: 41 subjects, 2-treatment crossover design, 6 equally-spaced
time points per period (2 periods), plus a baseline covariate.
492 observations total.

```{r}
data(DataSet17.1)
str(DataSet17.1)
cat("Subjects per sequence:\n")
print(table(unique(DataSet17.1[, c("id", "sequence")])$sequence))
```

```{r}
#| label: fig-17-1-spaghetti
#| fig-cap: "Individual trajectories by treatment and period"
ggplot(DataSet17.1,
       aes(x = t, y = y, group = id, colour = trt)) +
  geom_line(alpha = 0.4) +
  stat_summary(aes(group = trt), fun = mean, geom = "line",
               linewidth = 1.5) +
  facet_wrap(~ period, labeller = label_both) +
  labs(title = "Chapter 17: Subject trajectories by treatment and period",
       x = "Time within period", y = "Response",
       colour = "Treatment") +
  theme_bw()
```

### Covariance model comparison: CS, AR(1), ARH(1)

```{r}
## CS: compound symmetry via random subject intercept
fit_cs <- nlme::lme(
  fixed   = y ~ period + trt / t + baseline,
  random  = ~ 1 | id,
  data    = DataSet17.1,
  method  = "REML"
)

## AR(1)
fit_ar1 <- nlme::lme(
  fixed       = y ~ period + trt / t + baseline,
  random      = list(id = nlme::pdDiag(~ t)),
  correlation = nlme::corAR1(form = ~ t | id / period),
  data        = DataSet17.1,
  method      = "REML"
)

## ARH(1): heterogeneous variances per time + AR(1)
fit_arh1 <- nlme::lme(
  fixed       = y ~ period + trt / t + baseline,
  random      = list(id = nlme::pdDiag(~ t)),
  correlation = nlme::corAR1(form = ~ t | id / period),
  weights     = nlme::varIdent(form = ~ 1 | t),
  data        = DataSet17.1,
  method      = "REML"
)

stats::AIC(fit_cs, fit_ar1, fit_arh1)
```

### Treatment contrasts (equal intercepts, equal slopes)

```{r}
emm_int <- emmeans::emmeans(fit_arh1, ~ trt, at = list(t = 0))
emmeans::contrast(emm_int, method = "pairwise")

emm_slp <- emmeans::emtrends(fit_arh1, ~ trt, var = "t")
emmeans::contrast(emm_slp, method = "pairwise")
```

## Example 17.2 — Sparse Longitudinal with SP(POW) (Section 17.3.2)

`DataSet17.2`: 41 subjects (17 on trt=1, 24 on trt=2) observed at up to
9 unequally-spaced times (0, 1, 2, 4, 8, 16, 32, 64, 128).
Only 101 of 369 possible observations are present (sparse).

```{r}
data(DataSet17.2)
str(DataSet17.2)
cat("Observations per treatment:\n")
print(table(DataSet17.2$trt))
cat("Unique times:", sort(unique(DataSet17.2$time)), "\n")
```

```{r}
#| label: fig-17-2-profiles
#| fig-cap: "Sparse longitudinal profiles by treatment"
ggplot(DataSet17.2,
       aes(x = time, y = y, group = subject, colour = trt)) +
  geom_line(alpha = 0.4) +
  scale_x_continuous(trans = "log1p",
                     breaks = c(0, 1, 2, 4, 8, 16, 32, 64, 128)) +
  labs(title = "Chapter 17: Sparse profiles by treatment (log1p time scale)",
       x = "Time (log1p scale)", y = "Response",
       colour = "Treatment") +
  theme_bw()
```

### SP(POW) model: spatial-power covariance for unequal times

```{r}
## SP(POW) approximated via corExp (exponential = continuous AR(1))
fit_sppow <- nlme::lme(
  fixed       = y ~ trt + time + trt:time,
  random      = ~ 1 + time | subject,
  correlation = nlme::corExp(form = ~ time | subject, nugget = FALSE),
  data        = DataSet17.2,
  method      = "REML",
  control     = nlme::lmeControl(opt = "optim", maxIter = 200)
)
summary(fit_sppow)
```

```{r}
emm_trt <- emmeans::emmeans(fit_sppow, ~ trt,
                              at = list(time = c(0, 1, 4, 16, 64)))
print(emm_trt)
```

## Key Takeaways

- Repeated measures require a covariance model; CS is often adequate
  but ARH(1) better represents decaying and heterogeneous correlation.
- SP(POW) generalises AR(1) to unequally-spaced and incomplete time grids.
- Use `AIC` to select among covariance structures (fit with REML for
  final model, ML for covariance comparison if needed).
- `nlme` offers greater flexibility for covariance structures than `lme4`.

## References

Stroup, W. W., Ptukhina, M., and Garai, S. (2024). *Generalized Linear Mixed Models: Modern Concepts, Methods and Applications* (2nd ed.). CRC Press.

Verbeke, G., & Molenberghs, G. (2000).
*Linear Mixed Models for Longitudinal Data*. Springer.
