---
title: "Chapter 6: Inference, Part I"
subtitle: "The Unified Framework for Generalised Linear Mixed Models"
author: "Muhammad Yaseen"
date: last-modified
vignette: >
  %\VignetteIndexEntry{Chapter 6: Inference, Part I}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: show
    df-print: kable
    fig-width: 7
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(modernGLMM)
library(lme4)
library(emmeans)
library(ggplot2)
```

## Overview

Chapter 6 unifies the models developed in earlier chapters under the
**Generalised Linear Mixed Model (GLMM)** framework.

The GLMM has three components:

1. **Random component**: $Y_i \mid \mathbf{b} \sim$ exponential family
   with mean $\mu_i$
2. **Systematic component** (linear predictor):
   $\eta_i = \mathbf{x}_i^\top\pmb{\beta} + \mathbf{z}_i^\top\mathbf{b}$
3. **Link function**: $g(\mu_i) = \eta_i$

### Common link functions

| Distribution | Link | $g(\mu)$ |
|---|---|---|
| Gaussian | Identity | $\mu$ |
| Binomial | Logit | $\log[\mu/(1-\mu)]$ |
| Poisson | Log | $\log(\mu)$ |
| Gamma | Log | $\log(\mu)$ |
| Beta | Logit | $\log[\mu/(1-\mu)]$ |

### Random effects structure

$$\mathbf{b} \sim \mathcal{N}(\mathbf{0}, \mathbf{G}(\pmb{\theta}))$$

The variance component vector $\pmb{\theta}$ is estimated by
REML (for Gaussian) or by maximum likelihood / Laplace approximation
(for non-Gaussian families).

## The lme4 Formula Language

The GLMM formula in `lme4` follows a compact syntax:

```r
response ~ fixed_effects + (random_slope | grouping_factor)
```

| Formula part | Meaning |
|---|---|
| `(1 \| g)` | Random intercept by group $g$ |
| `(x \| g)` | Random intercept + slope for $x$ by $g$ |
| `(0 + x \| g)` | Random slope only (no intercept) |
| `(1 \| g/h)` | Random intercept, $h$ nested in $g$ |

## Illustration: Poisson GLMM

The following illustrates a Poisson GLMM using the blocked count data
from Chapter 11 (`DataSet11.3`):

```{r}
data(DataSet11.3)
DataSet11.3$block <- factor(DataSet11.3$block)
DataSet11.3$trt   <- factor(DataSet11.3$trt)

## Fit Poisson GLMM
fit_pois <- lme4::glmer(
  count ~ trt + (1 | block),
  family  = stats::poisson(link = "log"),
  data    = DataSet11.3,
  control = lme4::glmerControl(optimizer = "bobyqa")
)
summary(fit_pois)
```

```{r}
## Overdispersion check
if (requireNamespace("performance", quietly = TRUE)) {
  performance::check_overdispersion(fit_pois)
}
```

```{r}
## DHARMa diagnostics
if (requireNamespace("DHARMa", quietly = TRUE)) {
  DHARMa::simulateResiduals(fit_pois, plot = TRUE)
}
```

```{r}
emm6 <- emmeans::emmeans(fit_pois, ~ trt, type = "response")
print(emm6)
```

```{r}
if (requireNamespace("report", quietly = TRUE)) {
  report::report(fit_pois)
}
```

## The Modern GLMM Ecosystem

| Package | Purpose |
|---|---|
| `lme4` | Core LMM/GLMM fitting |
| `lmerTest` | Satterthwaite df for LMMs |
| `glmmTMB` | Flexible GLMMs (NB, beta, zero-inflated, etc.) |
| `emmeans` | Marginal means and contrasts |
| `DHARMa` | Simulation-based residual diagnostics |
| `performance` | ICC, overdispersion, model fit |
| `parameters` | Tidy parameter tables |
| `report` | Automatic model interpretation |
| `broom.mixed` | Tidy outputs |

## Key Takeaways

- The GLMM unifies all models: LM, GLM, LMM, and GLMM are special cases.
- The choice of link function and distribution is driven by the
  **data generating process**, not convenience.
- Modern tools (`DHARMa`, `performance`) provide powerful diagnostics
  unavailable in the original SAS/PROC GLIMMIX workflow.

## References

Stroup, W. W., Ptukhina, M., and Garai, S. (2024). *Generalized Linear Mixed Models: Modern Concepts, Methods and Applications* (2nd ed.). CRC Press.

Bates, D., Maechler, M., Bolker, B., & Walker, S. (2015). Fitting linear
mixed-effects models using lme4. *Journal of Statistical Software*, 67(1), 1–48.
