---
title: "Chapter 8: Treatment and Explanatory Variable Structure"
subtitle: "Factorial Designs, Interaction Models, and Response Surface Methods"
author: "Muhammad Yaseen"
date: last-modified
vignette: >
  %\VignetteIndexEntry{Chapter 8: Treatment and Explanatory Variable Structure}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: show
    df-print: kable
    fig-width: 7
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(modernGLMM)
library(lme4)
library(lmerTest)
library(emmeans)
library(ggplot2)
```

## Overview

Chapter 8 covers treatment and regression structures for GLMMs:

- **Factorial designs** with qualitative factors
- **Analysis of covariance** (ANCOVA)
- **Response surface methodology** (RSM)
- **Nested and crossed** factor structures

## Example 8.1 — Two-Way Factorial (All Qualitative)

The two-way factorial model:

$$y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \varepsilon_{ijk}$$

```{r}
data(DataSet8.1)
DataSet8.1$a <- factor(DataSet8.1$a)
DataSet8.1$b <- factor(DataSet8.1$b)
str(DataSet8.1)
```

```{r}
Exam8.1.lm <- stats::lm(y ~ a * b, data = DataSet8.1)
summary(Exam8.1.lm)
anova(Exam8.1.lm)
```

```{r}
emm8.1 <- emmeans::emmeans(Exam8.1.lm, ~ a * b)
print(emm8.1)
```

```{r}
## Simple effects: b within each level of a
emmeans::contrast(
  emmeans::emmeans(Exam8.1.lm, ~ b | a),
  method = "pairwise", adjust = "tukey"
)
```

```{r}
#| label: fig-8-1
#| fig-cap: "Interaction plot: Factor A × Factor B"
emmeans::emmip(Exam8.1.lm, a ~ b, CIs = TRUE) +
  theme_bw() +
  labs(title = "Chapter 8: A × B Interaction")
```

## Example 8.2 — ANCOVA

Analysis of covariance with treatment and continuous covariate $x$:

$$y_{ij} = \mu + \tau_i + \beta x_{ij} + \varepsilon_{ij}$$

```{r}
data(DataSet8.2)
DataSet8.2$trt <- factor(DataSet8.2$trt)
str(DataSet8.2)
```

```{r}
Exam8.2.lm <- stats::lm(y ~ trt + x, data = DataSet8.2)
summary(Exam8.2.lm)
anova(Exam8.2.lm)
```

```{r}
## Adjusted means (covariate at its mean)
emm8.2 <- emmeans::emmeans(Exam8.2.lm, ~ trt,
                            at = list(x = mean(DataSet8.2$x)))
print(emm8.2)
emmeans::contrast(emm8.2, method = "pairwise")
```

## Example 8.3 — Heterogeneous Slopes ANCOVA

```{r}
data(DataSet8.3)
DataSet8.3$trt <- factor(DataSet8.3$trt)

Exam8.3.lm <- stats::lm(y ~ trt * x, data = DataSet8.3)
summary(Exam8.3.lm)
anova(Exam8.3.lm)
```

```{r}
#| label: fig-8-3
#| fig-cap: "Heterogeneous slopes: y vs x by treatment"
ggplot(DataSet8.3, aes(x = x, y = y, colour = trt, group = trt)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Chapter 8: ANCOVA — Heterogeneous Slopes",
       x = "Covariate x", y = "Response y") +
  theme_bw()
```

## Example 8.7 — Response Surface

```{r}
data(DataSet8.7)
DataSet8.7$a <- factor(DataSet8.7$a)
str(DataSet8.7)
```

```{r}
## Quadratic response surface within groups
Exam8.7.lm <- stats::lm(y ~ a + x + I(x^2) + a:x, data = DataSet8.7)
summary(Exam8.7.lm)
```

```{r}
#| label: fig-8-7
#| fig-cap: "Response surface: y vs x by group a"
xseq <- seq(min(DataSet8.7$x), max(DataSet8.7$x), length.out = 50L)
pred_df <- do.call(rbind, lapply(levels(DataSet8.7$a), function(ai) {
  nd <- data.frame(a = factor(ai, levels = levels(DataSet8.7$a)), x = xseq)
  nd$y_hat <- stats::predict(Exam8.7.lm, newdata = nd)
  nd
}))

ggplot(DataSet8.7, aes(x = x, y = y, colour = a)) +
  geom_point(alpha = 0.5) +
  geom_line(data = pred_df, aes(x = x, y = y_hat, colour = a),
            linewidth = 1) +
  labs(title = "Chapter 8: Response surface by group",
       x = "x", y = "y", colour = "Group") +
  theme_bw()
```

## Key Takeaways

- Factorial interactions require **simple effects** comparisons
  (comparing one factor at each level of the other).
- ANCOVA requires checking the **homogeneity of slopes** assumption.
- `emmeans` handles adjusted means, custom contrasts, and
  interaction decomposition seamlessly.
- Response surface models use polynomial terms; for GLMMs, the
  polynomial is on the linear predictor scale.

## References

Stroup, W. W., Ptukhina, M., and Garai, S. (2024). *Generalized Linear Mixed Models: Modern Concepts, Methods and Applications* (2nd ed.). CRC Press.
