---
title: "Chapter 5: GLMM Estimation"
subtitle: "Exponential Family, ML, REML, and Quasi-Likelihood"
author: "Muhammad Yaseen"
date: last-modified
vignette: >
  %\VignetteIndexEntry{Chapter 5: GLMM Estimation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    df-print: kable
    fig-width: 7
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

## Overview

Chapter 5 of Stroup, Ptukhina & Garai (2024) presents the **estimation
framework for GLMMs** — the mathematical machinery behind likelihood-based
fitting of mixed models.  This chapter is entirely theoretical and contains
no applied examples or datasets.  Core topics are:

- Exponential family distributions and their canonical links
- Maximum Likelihood (ML) and Restricted Maximum Likelihood (REML)
- Newton-Raphson and Fisher scoring algorithms
- Quasi-likelihood and pseudo-likelihood
- Laplace and adaptive Gauss-Hermite approximations

## Exponential Family Distributions

A distribution belongs to the **exponential family** if its density can
be written as:

$$f(y;\,\theta, \phi) = \exp\!\left\{\frac{y\theta - b(\theta)}{a(\phi)} + c(y, \phi)\right\}$$

where $\theta$ is the **natural (canonical) parameter**, $b(\theta)$
determines the mean via $\mu = b'(\theta)$, and $\phi$ is a dispersion
parameter.

Key members relevant to GLMMs:

| Family | Link | $b(\theta)$ | Mean $\mu$ |
|--------|------|-------------|------------|
| Gaussian | identity | $\theta^2/2$ | $\theta$ |
| Binomial | logit | $\log(1 + e^\theta)$ | $e^\theta/(1+e^\theta)$ |
| Poisson | log | $e^\theta$ | $e^\theta$ |
| Gamma | reciprocal | $-\log(-\theta)$ | $-1/\theta$ |

```{r}
## Canonical link and variance function for common families
fams <- list(
  Gaussian  = stats::gaussian(link = "identity"),
  Binomial  = stats::binomial(link = "logit"),
  Poisson   = stats::poisson(link = "log"),
  Gamma     = stats::Gamma(link = "inverse")
)
cat("Family    | Link\n")
cat("----------|---------\n")
for (nm in names(fams)) cat(sprintf("%-10s| %s\n", nm, fams[[nm]]$link))
```

## Maximum Likelihood Estimation

The ML estimate maximises the log-likelihood $\ell(\boldsymbol{\theta}; \mathbf{y})$.
For a GLMM with fixed effects $\boldsymbol{\beta}$ and variance components
$\boldsymbol{\psi}$, the marginal log-likelihood integrates over the
random effects $\mathbf{b}$:

$$\ell(\boldsymbol{\beta}, \boldsymbol{\psi}) = \log \int f(\mathbf{y}\mid\mathbf{b};\boldsymbol{\beta})\,
p(\mathbf{b};\boldsymbol{\psi})\,d\mathbf{b}$$

This integral is intractable in general and requires approximation
(Laplace, AGHQ) or pseudo-likelihood methods.

## Restricted Maximum Likelihood (REML)

REML removes the contribution of $\boldsymbol{\beta}$ from the
likelihood before estimating $\boldsymbol{\psi}$.  For Gaussian LMMs
this is equivalent to maximising:

$$\ell_R(\boldsymbol{\psi}) = -\tfrac{1}{2}\log|\mathbf{V}|
  -\tfrac{1}{2}\log|\mathbf{X}^\top\mathbf{V}^{-1}\mathbf{X}|
  -\tfrac{1}{2}(\mathbf{y}-\mathbf{X}\hat{\boldsymbol{\beta}})^\top
    \mathbf{V}^{-1}(\mathbf{y}-\mathbf{X}\hat{\boldsymbol{\beta}})$$

REML produces **less biased** variance component estimates than ML
because it accounts for the degrees of freedom used by
$\hat{\boldsymbol{\beta}}$.  Use REML for final estimates; use ML when
comparing models with different fixed effects via likelihood ratio tests.

```{r}
## Illustration: ML vs REML variance component estimates
if (requireNamespace("lme4", quietly = TRUE)) {
  set.seed(42)
  n <- 30; nblk <- 6
  block <- rep(1:nblk, each = n / nblk)
  y     <- 5 + rnorm(nblk, sd = 2)[block] + rnorm(n)
  d     <- data.frame(y = y, block = factor(block))

  fit_ml   <- lme4::lmer(y ~ 1 + (1 | block), data = d, REML = FALSE)
  fit_reml <- lme4::lmer(y ~ 1 + (1 | block), data = d, REML = TRUE)

  vc_ml   <- as.data.frame(lme4::VarCorr(fit_ml))
  vc_reml <- as.data.frame(lme4::VarCorr(fit_reml))

  cat(sprintf("Block variance  — ML: %.4f  |  REML: %.4f\n",
              vc_ml$vcov[1], vc_reml$vcov[1]))
  cat(sprintf("Residual variance— ML: %.4f  |  REML: %.4f\n",
              vc_ml$vcov[2], vc_reml$vcov[2]))
}
```

## Newton-Raphson and Fisher Scoring

Both algorithms iterate:

$$\boldsymbol{\theta}^{(t+1)} = \boldsymbol{\theta}^{(t)} +
  \mathbf{H}^{-1}(\boldsymbol{\theta}^{(t)})\,
  \mathbf{s}(\boldsymbol{\theta}^{(t)})$$

where $\mathbf{s} = \partial\ell/\partial\boldsymbol{\theta}$ is the
score and $\mathbf{H}$ is the negative Hessian.  Fisher scoring replaces
$\mathbf{H}$ with the **expected information** matrix, guaranteeing
positive-definiteness and simplifying computation for GLMs.

## Quasi-Likelihood

When the full likelihood is unknown or intractable, **quasi-likelihood**
uses only the mean-variance relationship $V(\mu)$ to construct estimating
equations identical in form to the score equations:

$$U(\boldsymbol{\beta}) = \mathbf{D}^\top \mathbf{V}^{-1}(\mathbf{y} - \boldsymbol{\mu}) = \mathbf{0}$$

Quasi-likelihood estimates of $\boldsymbol{\beta}$ are consistent and
asymptotically normal under weak moment assumptions.  The dispersion
parameter $\phi$ is estimated by the Pearson statistic divided by
residual degrees of freedom.

## Laplace Approximation

For non-Gaussian GLMMs, the marginal likelihood is approximated at the
mode $\hat{\mathbf{b}}$ of the integrand:

$$\ell(\boldsymbol{\beta},\boldsymbol{\psi}) \approx \ell(\boldsymbol{\beta}, \hat{\mathbf{b}}, \boldsymbol{\psi})
  - \tfrac{1}{2}\log\left|{\mathbf{H}_{\hat{\mathbf{b}}}}\right| + \text{const}$$

Laplace is exact for Gaussian random effects when the conditional
distribution is also Gaussian (i.e., the Gaussian LMM).  For other
GLMMs it introduces $O(n^{-1})$ approximation error.  Adaptive
Gauss-Hermite quadrature (AGHQ) achieves higher accuracy at greater
computational cost.

## Key Takeaways

- ML underestimates variance components; REML corrects for fixed-effects
  df and is preferred for final reporting.
- Pseudo-likelihood (PL) used by SAS PROC GLIMMIX is a linearised
  approximation; `lme4` uses Laplace or AGHQ, which differ from PL and
  can yield slightly different point estimates.
- Chapter 5 has no applied examples; the estimation methods are
  demonstrated in practice throughout Chapters 8–21.

## References

Stroup, W. W., Ptukhina, M., and Garai, S. (2024). *Generalized Linear Mixed
Models: Modern Concepts, Methods and Applications* (2nd ed.). CRC Press.
Chapter 5: GLMM Estimation, pp. 146+.
