---
title: "Chapter 1: Modeling Basics"
subtitle: "Response Distributions, Linear Models, and Generalized Linear Models"
author: "Muhammad Yaseen"
date: last-modified
vignette: >
  %\VignetteIndexEntry{Chapter 1: Modeling Basics}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    theme: cosmo
    df-print: kable
    fig-width: 7
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(modernGLMM)
library(ggplot2)
library(emmeans)
if (requireNamespace("parameters", quietly = TRUE)) library(parameters)
```

## Overview

Chapter 1 of Stroup, Ptukhina, and Garai (2024) introduces the statistical landscape motivating
GLMMs.  The key insight is that different response types — Gaussian,
binomial, Poisson — require different probability models, and forcing all of
them into a normal-error linear model leads to inferential distortions.

The chapter uses a dose–response dataset (Table 1.1) to contrast:

1. A **linear model** (LM) — forced Gaussian assumption
2. A **generalized linear model** (GLM) with logistic link — appropriate
   binomial model

## Data

Table 1.1 records the number of successes $y$ out of $N_x$ trials across
11 dose levels $x = 0, 1, \ldots, 10$.

```{r}
data(Table1.1)
knitr::kable(Table1.1, caption = "Table 1.1: Dose-response data")
```

The observed proportion $p = y / N_x$ and the two candidate model
predictions are shown in Figure 1.1 below.

## Linear Model

The LM fits a straight line on the probability scale:

$$p_i = \beta_0 + \beta_1 x_i + \varepsilon_i, \quad \varepsilon_i \sim \mathcal{N}(0, \sigma^2)$$

```{r}
Exam1.1.lm <- stats::lm(formula = y / Nx ~ x, data = Table1.1)
if (requireNamespace("parameters", quietly = TRUE)) {
  parameters::model_parameters(Exam1.1.lm)
}
```

**Problem**: the fitted line can exceed $[0, 1]$ and assumes constant
variance — neither of which is appropriate for proportions.

## Generalized Linear Model (Logistic)

The logistic GLM uses:

$$\text{logit}(p_i) = \log\frac{p_i}{1 - p_i} = \beta_0 + \beta_1 x_i$$

so fitted values are automatically constrained to $(0, 1)$.

```{r}
Exam1.1.glm <- stats::glm(
  formula = cbind(y, Nx - y) ~ x,
  family  = stats::binomial(link = "logit"),
  data    = Table1.1
)
if (requireNamespace("parameters", quietly = TRUE)) {
  parameters::model_parameters(Exam1.1.glm)
}
```

```{r}
if (requireNamespace("report", quietly = TRUE)) {
  tryCatch(
    print(report::report(Exam1.1.glm)),
    error = function(e) message("report() not supported for this model: ", conditionMessage(e))
  )
}
```

## Comparison: LM vs GLM

```{r}
#| label: fig-lm-glm
#| fig-cap: "Figure 1.1 — LM (blue) versus logistic GLM (red) fitted proportions"
Table1.1$p_obs <- Table1.1$y / Table1.1$Nx
Table1.1$p_lm  <- Exam1.1.lm$fitted.values
Table1.1$p_glm <- Exam1.1.glm$fitted.values

plot_df <- rbind(
  data.frame(x = Table1.1$x, p = Table1.1$p_obs, Model = "Observed"),
  data.frame(x = Table1.1$x, p = Table1.1$p_lm,  Model = "LM"),
  data.frame(x = Table1.1$x, p = Table1.1$p_glm, Model = "GLM (logit)")
)

ggplot(plot_df, aes(x = x, y = p, colour = Model, shape = Model)) +
  geom_point(size = 2.5) +
  geom_line(data = subset(plot_df, Model != "Observed")) +
  scale_colour_manual(values = c("Observed" = "black",
                                 "LM"       = "#377EB8",
                                 "GLM (logit)" = "#E41A1C")) +
  labs(
    title = "Chapter 1: LM vs GLM for proportions",
    x     = "Dose (x)",
    y     = "Estimated proportion p",
    colour = "Model",
    shape  = "Model"
  ) +
  theme_bw()
```

## Correlation of Fitted Values with Observed

```{r}
cat("LM  correlation with observed:", round(cor(Table1.1$p_obs, Table1.1$p_lm),  4), "\n")
cat("GLM correlation with observed:", round(cor(Table1.1$p_obs, Table1.1$p_glm), 4), "\n")
```

The GLM consistently achieves higher correlation because it uses the
correct probability model.

## Estimated Marginal Means

```{r}
emm <- emmeans::emmeans(Exam1.1.glm, ~ 1, type = "response")
print(emm)
```

## Key Takeaways

- **Always match the probability distribution to the response type.**
- Linear models impose Gaussian assumptions that are inappropriate for
  bounded or count responses.
- GLMs generalise LMs through the link function and variance function.
- The logistic GLM is preferred for binomial proportions.

## References

Stroup, W. W., Ptukhina, M., and Garai, S. (2024). *Generalized Linear Mixed Models: Modern Concepts, Methods and Applications* (2nd ed.). CRC Press.
