---
title: "Chapter 20: Five Bayesian GLMM Examples"
subtitle: "Five Applied Bayesian GLMM Analyses"
author: "Muhammad Yaseen"
date: today
vignette: >
  %\VignetteIndexEntry{Chapter 20: Five Bayesian GLMM Examples}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: show
    df-print: kable
    fig-width: 7
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(modernGLMM)
library(lme4)
library(emmeans)
library(ggplot2)
```

## Overview

Chapter 20 presents five applied Bayesian GLMM analyses that parallel the
frequentist examples from earlier chapters. The analyses demonstrate:

1. **Gaussian LMM** (BLUP, Chapter 10) — Bayesian one-way random effects
2. **Poisson GLMM** (Count data, Chapter 11) — Bayesian insect count model
3. **Binomial GLMM** — Bayesian proportion analysis
4. **Ordinal GLMM** (Multinomial, Chapter 14) — Bayesian cumulative link model
5. **Repeated measures LMM** (Chapter 17) — Bayesian CS covariance structure

All examples use `brms` with weakly informative priors.

## Example 20.1 — Bayesian One-Way Random Effects (Gaussian)

```{r}
if (requireNamespace("brms", quietly = TRUE)) {
  data(DataSet10.1)
  DataSet10.1$a <- factor(DataSet10.1$a)

  fit20.1 <- brms::brm(
    y ~ 1 + (1 | a),
    data   = DataSet10.1,
    prior  = brms::prior(normal(0, 10), class = Intercept) +
             brms::prior(cauchy(0, 2.5), class = sd) +
             brms::prior(exponential(1), class = sigma),
    chains = 4, iter = 2000, seed = 20,
    silent = 2
  )
  print(summary(fit20.1))
  brms::pp_check(fit20.1, ndraws = 100)
}
```

## Example 20.2 — Bayesian Poisson GLMM (Count Data)

```{r}
if (requireNamespace("brms", quietly = TRUE)) {
  # Simulated count data
  set.seed(20)
  count_ex <- data.frame(
    trt   = factor(rep(c("A","B","C"), each = 12)),
    block = factor(rep(1:4, 9)),
    count = c(rpois(12, 5), rpois(12, 10), rpois(12, 15))
  )

  fit20.2 <- brms::brm(
    count ~ trt + (1 | block),
    data   = count_ex,
    family = poisson(),
    prior  = brms::prior(normal(0, 2), class = b) +
             brms::prior(normal(2, 1), class = Intercept) +
             brms::prior(cauchy(0, 1), class = sd),
    chains = 4, iter = 2000, seed = 20,
    silent = 2
  )
  print(emmeans::emmeans(fit20.2, ~ trt, type = "response"))
}
```

## Example 20.3 — Bayesian Binomial GLMM (Proportions)

```{r}
if (requireNamespace("brms", quietly = TRUE)) {
  data(DataSet12.2)
  DataSet12.2$a     <- factor(DataSet12.2$a)
  DataSet12.2$b     <- factor(DataSet12.2$b)
  DataSet12.2$block <- factor(DataSet12.2$block)

  fit20.3 <- brms::brm(
    f | trials(n) ~ a * b + (1 | block),
    data   = DataSet12.2,
    family = binomial(),
    prior  = brms::prior(normal(0, 5), class = b) +
             brms::prior(cauchy(0, 2), class = sd),
    chains = 4, iter = 2000, seed = 20,
    silent = 2
  )
  summary(fit20.3)
}
```

## Example 20.4 — Bayesian Ordinal GLMM

```{r}
if (requireNamespace("brms", quietly = TRUE)) {
  data(DataSet14.1)
  DataSet14.1$trt    <- factor(DataSet14.1$trt)
  DataSet14.1$blk    <- factor(DataSet14.1$blk)
  DataSet14.1$rating <- ordered(DataSet14.1$rating,
                                levels = c("slight", "modrat", "severe"))

  fit20.4 <- brms::brm(
    rating ~ trt + (1 | blk),
    data   = DataSet14.1,
    family = brms::cumulative("probit"),
    prior  = brms::prior(normal(0, 5), class = b) +
             brms::prior(cauchy(0, 2), class = sd),
    chains = 4, iter = 2000, seed = 20,
    silent = 2
  )
  summary(fit20.4)
}
```

## Example 20.5 — Bayesian Repeated Measures LMM

```{r}
if (requireNamespace("brms", quietly = TRUE)) {
  data(DataSet17.1)
  DataSet17.1$id  <- factor(DataSet17.1$id)
  DataSet17.1$trt <- factor(DataSet17.1$trt)
  DataSet17.1$t   <- factor(DataSet17.1$t)

  fit20.5 <- brms::brm(
    y ~ trt * t + (1 | id),
    data   = DataSet17.1,
    prior  = brms::prior(normal(0, 10), class = b) +
             brms::prior(cauchy(0, 5), class = sd) +
             brms::prior(exponential(1), class = sigma),
    chains = 4, iter = 2000, seed = 20,
    silent = 2
  )
  summary(fit20.5)
  emmeans::emmeans(fit20.5, ~ trt | t)
}
```

## Key Takeaways

- The `brms` formula syntax mirrors `lme4`, making Bayesian conversion
  straightforward.
- Prior specification should reflect domain knowledge; weakly informative
  defaults are safe starting points.
- `emmeans` works with `brms` objects for posterior marginal means.
- Posterior predictive checks (`brms::pp_check`) verify model fit.

## References

Stroup, W. W., Ptukhina, M., and Garai, S. (2024). *Generalized Linear Mixed
Models: Modern Concepts, Methods and Applications (2nd ed.)*. CRC Press.
