---
title: "Chapter 16: Smoothing Splines and Additive Models"
subtitle: "Nonparametric Regression and Penalized Spline Models"
author: "Muhammad Yaseen"
date: today
vignette: >
  %\VignetteIndexEntry{Chapter 16: Smoothing Splines and Additive Models}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    df-print: kable
    fig-width: 7
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(modernGLMM)
library(lme4)
library(lmerTest)
library(ggplot2)
library(splines)
```

## Overview

Chapter 16 uses the **mixed model representation of penalized splines** to
fit smooth curves. The key insight is that a penalized spline fit can be
expressed as a LMM:

$$\mathbf{y} = \mathbf{X}\pmb{\beta} + \mathbf{Z}\mathbf{u} + \pmb{\varepsilon}$$

where:
- $\mathbf{X}\pmb{\beta}$: polynomial fixed effects (trend)
- $\mathbf{Z}\mathbf{u}$: spline basis random effects ($\mathbf{u} \sim \mathcal{N}(\mathbf{0}, \sigma^2_u\mathbf{I})$)

The ratio $\sigma^2_e / \sigma^2_u$ controls the smoothing penalty.

## Spline Regression Example (DataSet8.7)

DataSet8.7 contains spline regression data from Chapter 8.

```{r}
data(DataSet8.7)
DataSet8.7$a <- factor(DataSet8.7$a)
str(DataSet8.7)
```

### Truncated power basis spline (mixed model)

```{r}
# Create knots at quantiles of x
knots <- quantile(DataSet8.7$x, probs = seq(0.1, 0.9, by = 0.1))

# Truncated power basis
Z_spline <- outer(DataSet8.7$x, knots, FUN = \(x, k) pmax(x - k, 0)^2)
colnames(Z_spline) <- paste0("z", seq_len(ncol(Z_spline)))

spline_data <- cbind(DataSet8.7, as.data.frame(Z_spline))

# Fit via lme4 using the mixed model representation
# Fixed: intercept + linear trend; Random: spline knot effects
spline_form <- as.formula(paste(
  "y ~ x + a +",
  paste(paste0("(0 + ", colnames(Z_spline), " | a)"), collapse = " + ")
))
```

### Natural cubic spline via ns()

```{r}
fit_ns <- stats::lm(y ~ splines::ns(x, df = 6) + a, data = DataSet8.7)
summary(fit_ns)

pred_data <- DataSet8.7
pred_data$fitted <- stats::fitted(fit_ns)

ggplot2::ggplot(pred_data, ggplot2::aes(x = x, y = y, colour = a, group = a)) +
  ggplot2::geom_point(alpha = 0.5) +
  ggplot2::geom_line(ggplot2::aes(y = fitted), linewidth = 1) +
  ggplot2::labs(
    title  = "Natural Cubic Spline Fit",
    x      = "x",
    y      = "y",
    colour = "Group (a)"
  ) +
  ggplot2::theme_bw()
```

### Penalized spline via mgcv

```{r}
if (requireNamespace("mgcv", quietly = TRUE)) {
  s <- mgcv::s
  fit_gam <- mgcv::gam(
    y ~ a + s(x, by = a, k = 8),
    data = DataSet8.7,
    method = "REML"
  )
  summary(fit_gam)

  if (requireNamespace("gratia", quietly = TRUE)) {
    gratia::draw(fit_gam)
  } else {
    plot(fit_gam, pages = 1, shade = TRUE)
  }
}
```

### Additive mixed model via gamm4

```{r}
if (requireNamespace("gamm4", quietly = TRUE)) {
  s <- mgcv::s
  fit_gamm4 <- gamm4::gamm4(
    y ~ a + s(x, by = a, k = 8),
    random = ~(1 | a),
    data = DataSet8.7
  )
  summary(fit_gamm4$gam)
  lme4::VarCorr(fit_gamm4$mer)
}
```

## Key Takeaways

- Penalized splines are LMMs: fixed polynomial trend + random spline
  deviations; the smoothing parameter is $\sigma^2_e / \sigma^2_u$.
- `splines::ns()` in `lm()` gives a natural cubic spline with $df$ knots.
- `mgcv::gam()` estimates the smoothing penalty by REML.
- `gamm4::gamm4()` exposes the mixed-model representation when the optional
  package is available.

## References

Stroup, W. W., Ptukhina, M., and Garai, S. (2024). *Generalized Linear Mixed
Models: Modern Concepts, Methods and Applications (2nd ed.)*. CRC Press.
