---
title: "Chapter 10: Best Linear Unbiased Prediction"
subtitle: "BLUP, Variance Components, and Random Effects Inference"
author: "Muhammad Yaseen"
date: last-modified
vignette: >
  %\VignetteIndexEntry{Chapter 10: Best Linear Unbiased Prediction}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: show
    df-print: kable
    fig-width: 7
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(modernGLMM)
library(lme4)
library(lmerTest)
library(emmeans)
```

## Overview

Chapter 10 introduces **Best Linear Unbiased Prediction (BLUP)** for random
effects in mixed models.  The key ideas are:

- **Variance decomposition**: total variance split into between-group
  ($\sigma^2_a$) and within-group ($\sigma^2_e$) components
- **BLUP shrinkage**: group-specific predictions are shrunk toward the overall
  mean — more so when sample sizes are small or when $\sigma^2_a / \sigma^2_e$ is small
- **Broad vs narrow inference**: broad inference averages over all possible
  random effect levels; narrow inference conditions on observed levels

The mixed model for observation $y_{ij}$ in group $i$, observation $j$:

$$y_{ij} = \mu + a_i + e_{ij}, \quad a_i \sim \mathcal{N}(0, \sigma^2_a),\quad e_{ij} \sim \mathcal{N}(0, \sigma^2_e)$$

## Example 10.1 — One-Way Random Effects Model

DataSet 10.1: 12 groups, 2 observations each (continuous response $y$).

```{r}
data(DataSet10.1)
DataSet10.1$a <- factor(DataSet10.1$a)
str(DataSet10.1)
```

### Random effects model

```{r}
Exam10.1Lmer <- lmerTest::lmer(
  y ~ 1 + (1 | a),
  data    = DataSet10.1,
  control = lme4::lmerControl(optimizer = "bobyqa")
)
summary(Exam10.1Lmer)
```

### Variance components

```{r}
as.data.frame(lme4::VarCorr(Exam10.1Lmer))
```

The intra-class correlation (ICC) quantifies the proportion of total
variance attributable to group differences:

$$\text{ICC} = \frac{\sigma^2_a}{\sigma^2_a + \sigma^2_e}$$

```{r}
if (requireNamespace("performance", quietly = TRUE)) {
  performance::icc(Exam10.1Lmer)
}
```

### BLUPs (Best Linear Unbiased Predictors)

```{r}
blup <- unlist(lme4::ranef(Exam10.1Lmer))
blup_df <- data.frame(
  group    = names(blup),
  intercept = blup,
  group_mean = mean(DataSet10.1$y) + blup
)
knitr::kable(blup_df, digits = 3,
             caption = "Table 10.1: BLUP estimates")
```

### Overall mean (narrow inference)

```{r}
emm10.1 <- emmeans::emmeans(Exam10.1Lmer, ~ 1)
print(emm10.1)
```

### Report

```{r}
if (requireNamespace("report", quietly = TRUE)) {
  report::report(Exam10.1Lmer)
}
```

## Example 10.2 — Two-Way Nested Random Effects Model

```{r}
data(DataSet10.2)
DataSet10.2$a <- factor(DataSet10.2$a)
DataSet10.2$b <- factor(DataSet10.2$b)
str(DataSet10.2)
```

```{r}
Exam10.2Lmer <- lmerTest::lmer(
  y ~ 1 + (1 | a / b),
  data    = DataSet10.2,
  control = lme4::lmerControl(optimizer = "bobyqa")
)
summary(Exam10.2Lmer)
```

```{r}
emm10.2 <- emmeans::emmeans(Exam10.2Lmer, ~ 1)
print(emm10.2)
```

## Example 10.4 — BLUP vs Fixed Effect Estimators

```{r}
data(DataSet10.4)
DataSet10.4$a <- factor(DataSet10.4$a)
DataSet10.4$b <- factor(DataSet10.4$b)
str(DataSet10.4)
```

```{r}
Exam10.4Lmer <- lmerTest::lmer(
  y ~ a + (1 | b) + (1 | b:a),
  data    = DataSet10.4,
  control = lme4::lmerControl(optimizer = "bobyqa")
)
stats::anova(Exam10.4Lmer, ddf = "Kenward-Roger")
emmeans::emmeans(Exam10.4Lmer, ~ a)
```

## Diagnostics

```{r}
if (requireNamespace("DHARMa", quietly = TRUE)) {
  sim_r <- DHARMa::simulateResiduals(Exam10.1Lmer, plot = TRUE)
}
```

## Key Takeaways

- Random effects models decompose variance into between-group ($\sigma^2_a$)
  and within-group ($\sigma^2_e$) components.
- BLUPs "shrink" group means toward the overall mean — particularly
  useful when group sample sizes are small.
- ICC quantifies clustering: high ICC → observations within groups are
  very similar.
- `emmeans` provides overall marginal means under both narrow and broad
  inference depending on the `lmer.df` option.

## References

Stroup, W. W., Ptukhina, M., and Garai, S. (2024). *Generalized Linear Mixed Models: Modern Concepts,
Methods and Applications*. CRC Press.
