---
title: "Chapter 3: Setting the Stage"
subtitle: "LM and GLM Methods for Fixed Effects"
author: "Muhammad Yaseen"
date: last-modified
vignette: >
  %\VignetteIndexEntry{Chapter 3: Setting the Stage}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    df-print: kable
    fig-width: 7
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(modernGLMM)
library(emmeans)
library(ggplot2)
```

## Overview

Chapter 3 covers fixed-effects analysis of designed experiments using
both linear models (LM) and generalised linear models (GLM).  The three
datasets illustrate:

- **DataSet3.1**: Two-treatment comparison with Gaussian and binomial responses
- **DataSet3.2**: Multi-location, multi-treatment factorial
- **DataSet3.3**: Time-course experiment with batch effects

## Example 3.2 — Two-Treatment Binomial GLM

The model is:

$$\text{logit}(p_i) = \mu + \tau_i, \quad i = 0, 1$$

```{r}
data(DataSet3.1)
DataSet3.1$trt <- factor(DataSet3.1$trt)
str(DataSet3.1)
```

```{r}
Exam3.2.glm <- stats::glm(
  cbind(F, N - F) ~ trt,
  family = stats::binomial(link = "logit"),
  data   = DataSet3.1
)
summary(Exam3.2.glm)
```

```{r}
if (requireNamespace("parameters", quietly = TRUE)) {
  parameters::model_parameters(Exam3.2.glm)
}
```

```{r}
## Marginal means on probability scale
emm3.2 <- emmeans::emmeans(Exam3.2.glm, ~ trt, type = "response")
print(emm3.2)
emmeans::contrast(emm3.2, method = "pairwise")
```

```{r}
if (requireNamespace("report", quietly = TRUE)) {
  tryCatch(
    print(report::report(Exam3.2.glm)),
    error = function(e) message("report() not supported for this model: ", conditionMessage(e))
  )
}
```

## Example 3.3 — Multi-Location Factorial

```{r}
data(DataSet3.2)
DataSet3.2$loc <- factor(DataSet3.2$loc)
DataSet3.2$trt <- factor(DataSet3.2$trt)
str(DataSet3.2)
```

```{r}
Exam3.3.lm <- stats::lm(Y ~ loc + trt, data = DataSet3.2)
summary(Exam3.3.lm)
anova(Exam3.3.lm)
```

```{r}
emm3.3 <- emmeans::emmeans(Exam3.3.lm, ~ trt)
print(emm3.3)
emmeans::contrast(emm3.3, method = "pairwise", adjust = "none")
```

### Interaction plot

```{r}
#| label: fig-3-3
#| fig-cap: "Treatment means averaged over locations"
emm_plot3.3 <- as.data.frame(emm3.3)
ggplot(emm_plot3.3, aes(x = trt, y = emmean)) +
  geom_point(size = 3, colour = "#2c5f8a") +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.12) +
  theme_bw() +
  labs(title = "Chapter 3: Treatment least-squares means",
       y = "Estimated marginal mean (Y)")
```

## Example 3.5 — Factorial Treatment Structure

```{r}
DataSet3.2$A <- factor(DataSet3.2$A)
DataSet3.2$B <- factor(DataSet3.2$B)
str(DataSet3.2)
```

```{r}
Exam3.5.lm <- stats::lm(Y ~ A * B + loc, data = DataSet3.2)
summary(Exam3.5.lm)
anova(Exam3.5.lm)
```

```{r}
emm3.5 <- emmeans::emmeans(Exam3.5.lm, ~ A * B)
print(emm3.5)
emmeans::contrast(
  emmeans::emmeans(Exam3.5.lm, ~ A | B),
  method = "pairwise",
  by = "B"
)
emmeans::contrast(
  emmeans::emmeans(Exam3.5.lm, ~ B | A),
  method = "pairwise",
  by = "A"
)
```

## Key Takeaways

- Use the binomial-count response `cbind(successes, failures)` when the
  book model is binomial with known denominators.
- `emmeans` provides estimated marginal means with proper
  standard errors accounting for the design structure.
- Interactions between treatment and location/batch should always be
  tested before interpreting main effects.

## References

Stroup, W. W., Ptukhina, M., and Garai, S. (2024). *Generalized Linear Mixed Models: Modern Concepts, Methods and Applications* (2nd ed.). CRC Press.
