---
title: "Chapter 13: Zero-Inflated and Hurdle Models"
subtitle: "Handling Excess Zeros in Count and Continuous Responses"
author: "Muhammad Yaseen"
date: today
vignette: >
  %\VignetteIndexEntry{Chapter 13: Zero-Inflated and Hurdle Models}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: [cosmo, theme.scss]
    df-print: kable
    fig-width: 7
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(modernGLMM)
library(emmeans)
```

## Overview

Chapter 13 addresses responses with **excess zeros** — more zeros than a
standard Poisson or binomial model predicts.

Two modelling strategies:

| Model | Structure | Interpretation |
|---|---|---|
| **Zero-Inflated (ZIP/ZINB)** | Mixture: structural zeros + count process | Some units can **never** have a non-zero count |
| **Hurdle** | Two-part: binary + truncated count | All non-zero counts come from a separate count process |

The zero-inflated Poisson model:

$$P(Y_i = 0) = \pi_i + (1 - \pi_i)e^{-\lambda_i}$$
$$P(Y_i = y) = (1 - \pi_i)\frac{e^{-\lambda_i}\lambda_i^y}{y!}, \quad y > 0$$

## Zero-Inflated Poisson GLMM

```{r}
if (requireNamespace("glmmTMB", quietly = TRUE)) {
  simulate_zinb <- function(n, mu, size, zprob) {
    structural_zero <- stats::rbinom(n, size = 1, prob = zprob) == 1
    counts <- stats::rnbinom(n, mu = mu, size = size)
    counts[structural_zero] <- 0L
    counts
  }

  set.seed(123)
  zi_data <- data.frame(
    trt = factor(rep(c("control", "low", "high"), each = 20)),
    block = factor(rep(1:5, times = 12)),
    count = c(
      simulate_zinb(20, mu = 2, size = 2, zprob = 0.5),
      simulate_zinb(20, mu = 5, size = 2, zprob = 0.3),
      simulate_zinb(20, mu = 10, size = 2, zprob = 0.1)
    )
  )

  # Zero-inflated Poisson GLMM
  fit_zip <- glmmTMB::glmmTMB(
    count ~ trt + (1 | block),
    ziformula = ~ trt,
    data   = zi_data,
    family = poisson()
  )
  summary(fit_zip)
}
```

## Hurdle Model

```{r}
if (requireNamespace("glmmTMB", quietly = TRUE) && exists("zi_data")) {
  fit_hurdle <- glmmTMB::glmmTMB(
    count ~ trt + (1 | block),
    ziformula = ~ trt,
    data   = zi_data,
    family = glmmTMB::truncated_poisson()
  )
  summary(fit_hurdle)
}
```

## Diagnostics

```{r}
if (requireNamespace("glmmTMB", quietly = TRUE) &&
    requireNamespace("DHARMa", quietly = TRUE) &&
    exists("fit_zip")) {
  sim_r <- DHARMa::simulateResiduals(fit_zip, plot = TRUE)
  DHARMa::testZeroInflation(sim_r)
}
```

## Key Takeaways

- Zero-inflated models are appropriate when zeros arise from two mechanisms:
  structural (always-zero) and sampling (could be non-zero).
- Hurdle models assume all zeros come from the binary part.
- `glmmTMB::glmmTMB()` fits both types; use `ziformula` to specify the
  zero-inflation component.
- `DHARMa::testZeroInflation()` tests whether residual zeros exceed
  what the fitted model predicts.

## References

Stroup, W. W., Ptukhina, M., and Garai, S. (2024). *Generalized Linear Mixed
Models: Modern Concepts, Methods and Applications (2nd ed.)*. CRC Press.
