---
title: "Chapter 2: Design Matters"
subtitle: "Rationale, Small-Sample Methods, and Two-Sample Comparisons"
author: "Muhammad Yaseen"
date: last-modified
vignette: >
  %\VignetteIndexEntry{Chapter 2: Design Matters}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    df-print: kable
    fig-width: 7
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(modernGLMM)
library(emmeans)
library(ggplot2)
```

## Overview

Chapter 2 establishes the inferential framework used throughout the book.
Key topics include:

- Likelihood-based inference
- Wald, likelihood ratio, and score tests
- Small-sample adjustments (Kenward-Roger, Satterthwaite)
- Two-sample design examples

## Example 2.B.2 — Dose-Response Binomial Data

```{r}
data(DataExam2.B.2)
str(DataExam2.B.2)
knitr::kable(DataExam2.B.2,
             caption = "Example 2.B.2: Dose-response binomial data")
```

### Logistic regression

```{r}
fit_2b2 <- stats::glm(
  cbind(y, n - y) ~ x,
  family = stats::binomial(link = "logit"),
  data   = DataExam2.B.2
)
summary(fit_2b2)
```

```{r}
if (requireNamespace("parameters", quietly = TRUE)) {
  parameters::model_parameters(fit_2b2)
}
```

```{r}
if (requireNamespace("report", quietly = TRUE)) {
  report::report(fit_2b2)
}
```

### Predicted probabilities

```{r}
#| label: fig-2b2-fit
#| fig-cap: "Fitted logistic curve for Example 2.B.2"
xnew <- data.frame(x = seq(0, 10, by = 0.1))
xnew$p_hat <- stats::predict(fit_2b2, newdata = xnew, type = "response")

ggplot() +
  geom_line(data = xnew, aes(x = x, y = p_hat), colour = "#E41A1C") +
  geom_point(data = DataExam2.B.2,
             aes(x = x, y = y / n), size = 2.5) +
  labs(title = "Example 2.B.2: Logistic dose-response",
       x = "Dose (x)", y = "Proportion (y/n)") +
  theme_bw()
```

## Example 2.B.3 — Three Treatment Comparison (Gaussian)

```{r}
data(DataExam2.B.3)
DataExam2.B.3$trt <- factor(DataExam2.B.3$trt)
str(DataExam2.B.3)
```

```{r}
fit_2b3 <- stats::lm(y ~ trt, data = DataExam2.B.3)
summary(fit_2b3)
anova(fit_2b3)
```

```{r}
emm_2b3 <- emmeans::emmeans(fit_2b3, ~ trt)
print(emm_2b3)
emmeans::contrast(emm_2b3, method = "pairwise", adjust = "tukey")
```

## Example 2.B.4 — Binomial Response, Three Treatments

```{r}
data(DataExam2.B.4)
DataExam2.B.4$trt <- factor(DataExam2.B.4$trt)
str(DataExam2.B.4)
```

```{r}
fit_2b4 <- stats::glm(
  cbind(Yij, Nij - Yij) ~ trt,
  family = stats::binomial(link = "logit"),
  data   = DataExam2.B.4
)
summary(fit_2b4)
```

```{r}
emm_2b4 <- emmeans::emmeans(fit_2b4, ~ trt, type = "response")
print(emm_2b4)
emmeans::contrast(emm_2b4, method = "pairwise")
```

## Example 2.B.7 — Two-Way Factorial

```{r}
data(DataExam2.B.7)
DataExam2.B.7$Rep <- factor(DataExam2.B.7$Rep)
DataExam2.B.7$a   <- factor(DataExam2.B.7$a)
DataExam2.B.7$b   <- factor(DataExam2.B.7$b)
str(DataExam2.B.7)
```

```{r}
fit_2b7 <- lmerTest::lmer(
  y ~ a * b + (1 | Rep),
  data    = DataExam2.B.7,
  control = lme4::lmerControl(optimizer = "bobyqa")
)
summary(fit_2b7)
anova(fit_2b7)
```

```{r}
emm_2b7 <- emmeans::emmeans(fit_2b7, ~ a * b)
emmeans::contrast(emm_2b7, method = "pairwise",
                  by = "a", adjust = "tukey")
```

## Key Takeaways

- Likelihood-based inference is the foundation of GLMMs.
- For Gaussian models, $t$- and $F$-tests follow exactly; for GLMMs,
  Wald and likelihood ratio statistics are approximate.
- `emmeans` provides Satterthwaite-adjusted degrees of freedom for
  mixed model comparisons.

## References

Stroup, W. W., Ptukhina, M., and Garai, S. (2024). *Generalized Linear Mixed Models: Modern Concepts, Methods and Applications* (2nd ed.). CRC Press.
