---
title: "Chapter 18: Correlated Errors, Part II: Spatial Variability"
subtitle: "Spatial Covariance Models for Gaussian and Non-Gaussian Responses"
author: "Muhammad Yaseen"
date: last-modified
vignette: >
  %\VignetteIndexEntry{Chapter 18: Correlated Errors, Part II: Spatial Variability}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    df-print: kable
    fig-width: 8
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(modernGLMM)
library(nlme)
library(lme4)
library(emmeans)
library(ggplot2)
```

## Overview

Chapter 18 addresses **spatial variability** where residual errors are
correlated as a function of distance between plots on a field.

### Spatial Covariance Functions

The spherical model (SP(SPH)):
$$C(d) = \sigma^2 \left[1 - \frac{3d}{2r} + \frac{d^3}{2r^3}\right]
         \mathbf{1}(d \le r)$$

The exponential model:
$$C(d) = \sigma^2 \exp\left(-\frac{d}{\phi}\right)$$

where $r$ or $\phi$ is the range parameter and $\sigma^2$ is the sill.

## Example 18.1 — Alliance Wheat Trial: Gaussian Spatial (Section 18.3)

`DataSet18.1`: 12×12 field grid, 48 wheat variety treatments, 3 complete
column blocks (columns 1–4, 5–8, 9–12). 144 plots total.

```{r}
data(DataSet18.1)
str(DataSet18.1)
with(DataSet18.1, table(block))
```

```{r}
#| label: fig-18-1-field
#| fig-cap: "Spatial yield map (12×12 grid)"
ggplot(DataSet18.1, aes(x = col, y = row, fill = y)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Chapter 18: Alliance wheat trial — yield map",
       x = "Column", y = "Row", fill = "Yield") +
  coord_equal() +
  theme_minimal()
```

### Baseline: complete block model (no spatial covariance)

```{r}
fit_rcb <- stats::lm(y ~ trt + block, data = DataSet18.1)
cat("RCB AIC:", stats::AIC(fit_rcb), "\n")
```

### Spatial models via nlme::gls

```{r}
## Spherical covariance (SP(SPH) — best model in book)
fit_sph <- nlme::gls(
  model       = y ~ trt + block,
  correlation = nlme::corSpher(form = ~ row + col, nugget = FALSE),
  data        = DataSet18.1,
  method      = "REML"
)

## Exponential covariance
fit_exp <- nlme::gls(
  model       = y ~ trt + block,
  correlation = nlme::corExp(form = ~ row + col, nugget = FALSE),
  data        = DataSet18.1,
  method      = "REML"
)

## AICC comparison (smallest = best)
stats::AIC(fit_sph, fit_exp)
```

### Spatial range and sill from best model

```{r}
tryCatch(
  nlme::intervals(fit_sph, which = "var-cov"),
  error = function(e) {
    cat("Intervals not available (non-PD Hessian on reconstructed data):",
        conditionMessage(e), "\n")
    cat("Range estimate (corSpher range parameter):\n")
    print(stats::coef(fit_sph$modelStruct$corStruct, unconstrained = FALSE))
  }
)
```

### Empirical variogram

```{r}
#| label: fig-18-1-vario
#| fig-cap: "Empirical variogram from spherical model"
vario <- nlme::Variogram(fit_sph, form = ~ row + col, resType = "normalized")
plot(vario, smooth = TRUE, main = "Chapter 18: Empirical variogram (spherical)")
```

## Example 18.2 — Hessian Fly Trial: Binomial Spatial (Section 18.4)

`DataSet18.2`: 16 wheat varieties in a 4×4 lattice design (4 complete
blocks), 64 plots. Response is number of Hessian fly-damaged plants
out of total (`y / n`).

```{r}
data(DataSet18.2)
str(DataSet18.2)
## Observed resistance rates by variety
with(DataSet18.2, tapply(y / n, variety, mean))
```

```{r}
#| label: fig-18-2-field
#| fig-cap: "Hessian fly damage map (spatial layout)"
ggplot(DataSet18.2, aes(x = col, y = row, fill = y / n)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  scale_fill_viridis_c(option = "inferno", direction = -1) +
  labs(title = "Chapter 18: Hessian fly damage rate map",
       x = "Column", y = "Row", fill = "Damage rate") +
  coord_equal() +
  theme_minimal()
```

### RCB model (baseline)

```{r}
fit_rcb18 <- lme4::glmer(
  cbind(y, n - y) ~ variety + (1 | block),
  family  = stats::binomial(link = "logit"),
  data    = DataSet18.2,
  control = lme4::glmerControl(optimizer = "bobyqa")
)
summary(fit_rcb18)
```

### G-side spherical spatial GLMM via glmmTMB

```{r}
if (requireNamespace("glmmTMB", quietly = TRUE)) {
  DataSet18.2$pos <- glmmTMB::numFactor(DataSet18.2$row, DataSet18.2$col)

  fit_sp18 <- glmmTMB::glmmTMB(
    cbind(y, n - y) ~ variety + (1 | block) +
      exp(pos + 0 | block),
    family = stats::binomial(link = "logit"),
    data   = DataSet18.2
  )
  summary(fit_sp18)

  emm_sp18 <- emmeans::emmeans(fit_sp18, ~ variety, type = "response")
  print(emm_sp18)
}
```

## Key Takeaways

- Ignoring spatial correlation leads to underestimated standard errors
  and inflated Type I error for treatment comparisons.
- `nlme::gls()` provides spherical, exponential, and Gaussian spatial
  covariance for Gaussian responses; select the best model by AICC.
- For non-Gaussian spatial GLMMs (e.g., binomial), use `glmmTMB` with
  spatial random effects.
- The variogram is the primary diagnostic for spatial dependence.

## References

Stroup, W. W., Ptukhina, M., and Garai, S. (2024). *Generalized Linear Mixed Models: Modern Concepts, Methods and Applications* (2nd ed.). CRC Press.

Pinheiro, J., & Bates, D. (2000).
*Mixed-Effects Models in S and S-PLUS*. Springer.
