---
title: "Chapter 15: Time-to-Event Data"
subtitle: "Survival Analysis and Frailty Models"
author: "Muhammad Yaseen"
date: today
vignette: >
  %\VignetteIndexEntry{Chapter 15: Time-to-Event Data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: [cosmo, theme.scss]
    df-print: kable
    fig-width: 7
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(modernGLMM)
library(ggplot2)
```

## Overview

Chapter 15 covers **time-to-event** (survival) data in the GLMM context.
A key distinction from standard survival analysis is the inclusion of
**random effects (frailty)** to account for clustering.

Core concepts:

- **Survival function**: $S(t) = P(T > t)$
- **Hazard function**: $h(t) = \lim_{\Delta t \to 0} P(t \le T < t + \Delta t \mid T \ge t) / \Delta t$
- **Cox PH model**: $h(t \mid \mathbf{x}) = h_0(t) \exp(\mathbf{x}^\top\boldsymbol{\beta})$
- **Frailty model**: adds a random effect $u_i$ on the log hazard scale

$$h(t \mid \mathbf{x}_i, u_i) = h_0(t) \exp(\mathbf{x}_i^\top\boldsymbol{\beta} + u_i), \quad u_i \sim \mathcal{N}(0, \sigma^2)$$

## Kaplan-Meier Curves

```{r}
if (requireNamespace("survival", quietly = TRUE)) {
  # Simulated survival data: 3 treatments, 4 clusters
  set.seed(99)
  n <- 60
  surv_data <- data.frame(
    id      = 1:n,
    time    = rexp(n, rate = c(rep(0.1, 20), rep(0.2, 20), rep(0.3, 20))),
    status  = rbinom(n, 1, 0.8),
    trt     = factor(rep(c("A","B","C"), each = 20)),
    cluster = factor(sample(1:4, n, replace = TRUE))
  )

  km_fit <- survival::survfit(
    survival::Surv(time, status) ~ trt,
    data = surv_data
  )

  if (requireNamespace("ggsurvfit", quietly = TRUE)) {
    ggsurvfit::ggsurvfit(km_fit) +
      ggsurvfit::add_confidence_interval() +
      ggsurvfit::add_risktable()
  } else {
    plot(km_fit, col = 1:3, lwd = 2,
         xlab = "Time", ylab = "Survival probability",
         main = "Kaplan-Meier Curves by Treatment")
    legend("topright", levels(surv_data$trt), col = 1:3, lwd = 2)
  }
}
```

## Cox Proportional Hazards Frailty Model

```{r}
if (requireNamespace("survival", quietly = TRUE) && exists("surv_data")) {
  frailty <- survival::frailty
  cox_frailty <- survival::coxph(
    survival::Surv(time, status) ~ trt + frailty(cluster),
    data = surv_data
  )
  stats::coef(cox_frailty)
}
```

## Cox Mixed-Effects Model

```{r}
if (requireNamespace("survival", quietly = TRUE) &&
    requireNamespace("coxme", quietly = TRUE) &&
    exists("surv_data")) {
  fit_coxme <- coxme::coxme(
    survival::Surv(time, status) ~ trt + (1 | cluster),
    data = surv_data
  )
  summary(fit_coxme)
}
```

## Key Takeaways

- Frailty models add cluster-level random effects to the Cox model —
  unobserved heterogeneity inflates the hazard ratio if ignored.
- `coxme::coxme()` fits Cox mixed-effects models when a Gaussian frailty
  term is needed.
- `survival::survfit()` computes Kaplan-Meier curves; `ggsurvfit` is used
  when available for ggplot2 output.

## References

Stroup, W. W., Ptukhina, M., and Garai, S. (2024). *Generalized Linear Mixed
Models: Modern Concepts, Methods and Applications (2nd ed.)*. CRC Press.
