---
title: "Chapter 14: Multinomial Data"
subtitle: "Ordinal and Nominal Categorical Responses in Mixed Models"
author: "Muhammad Yaseen"
date: last-modified
vignette: >
  %\VignetteIndexEntry{Chapter 14: Multinomial Data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    df-print: kable
    fig-width: 7
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(modernGLMM)
library(emmeans)
library(ggplot2)
```

## Overview

Chapter 14 extends GLMMs to **categorical responses with more than two
categories**.  Two cases arise:

1. **Ordinal** (ordered categories): $Y \in \{1 < 2 < \cdots < J\}$
2. **Nominal** (unordered categories): $Y \in \{1, 2, \ldots, J\}$

## Ordinal Models

### Proportional-Odds Model (Cumulative Logit)

$$\text{logit}[P(Y \le j \mid x)] = \theta_j - \mathbf{x}^\top\boldsymbol{\beta}$$

- $\theta_1 < \theta_2 < \cdots < \theta_{J-1}$: threshold intercepts
- $\boldsymbol{\beta}$: common slope vector (proportional odds assumption)
- The minus sign convention means larger $\beta$ → higher probability of
  higher categories

## Example 14.1 — Ordinal GLMM (Section 14.3)

`DataSet14.1`: 10 blocks × 6 treatments × 3 ordered ratings
(slight < modrat < severe), stored as frequency counts (`y`).

```{r}
data(DataSet14.1)
str(DataSet14.1)
## Marginal rating distribution by treatment
with(DataSet14.1, tapply(y, list(trt, rating), sum))
```

```{r}
#| label: fig-14-1-bar
#| fig-cap: "Distribution of ordinal ratings by treatment (marginal counts)"
marg14 <- aggregate(y ~ trt + rating, data = DataSet14.1, FUN = sum)
marg14$rating <- factor(marg14$rating,
                        levels = c("slight", "modrat", "severe"),
                        ordered = TRUE)
ggplot(marg14, aes(x = trt, y = y, fill = rating)) +
  geom_col(position = "fill") +
  labs(title = "Chapter 14: Ordinal rating distributions by treatment",
       x = "Treatment", y = "Proportion",
       fill = "Rating") +
  scale_fill_brewer(palette = "Blues") +
  theme_bw()
```

```{r}
## Fixed-effects proportional-odds via MASS::polr (no random block effect)
## NOTE: ignores block random effect; for illustration of fixed-effect structure.
## A full proportional-odds GLMM would require ordinal::clmm or glmmTMB >= 1.2.
if (requireNamespace("MASS", quietly = TRUE)) {
  fit_polr <- MASS::polr(
    rating ~ trt,
    weights = y,
    data    = DataSet14.1,
    Hess    = TRUE,
    method  = "logistic"
  )
  summary(fit_polr)

  ## Wald p-values
  coef_tab <- coef(summary(fit_polr))
  pval     <- stats::pnorm(abs(coef_tab[, "t value"]),
                            lower.tail = FALSE) * 2
  cbind(coef_tab, "p value" = round(pval, 4))

  ## Odds ratios for treatments vs reference (trt 5)
  exp(coef(fit_polr)[seq_len(nlevels(DataSet14.1$trt) - 1L)])
}
```

## Non-Proportional Odds (Section 14.4)

### Baseline-Category Logit

For unordered categories with category $J$ as the reference:

$$\log\frac{P(Y = j)}{P(Y = J)} = \alpha_j + \tau_{ij}, \quad j = 1,\ldots,J-1$$

Each non-reference category has its own intercept and slope vector.

## Example 14.2 — Non-Proportional Odds: Poinsettia Trial

`DataSet14.2`: 12 growers × 3 varieties × 3 quality ratings
(poor < average < good), stored as frequency counts (`y`).
The proportional-odds assumption fails; variety effects differ by
rating boundary.

```{r}
data(DataSet14.2)
str(DataSet14.2)
## Marginal variety × rating totals (match published table p.438)
with(DataSet14.2, tapply(y, list(variety, rating), sum))
```

```{r}
#| label: fig-14-2-bar
#| fig-cap: "Quality rating distribution by variety (marginal counts)"
marg14b <- aggregate(y ~ variety + rating, data = DataSet14.2, FUN = sum)
marg14b$rating <- factor(marg14b$rating, levels = c("A", "B", "C"))
ggplot(marg14b, aes(x = variety, y = y, fill = rating)) +
  geom_col(position = "fill") +
  labs(title = "Chapter 14: Poinsettia quality by variety",
       x = "Variety", y = "Proportion",
       fill = "Rating") +
  scale_fill_brewer(palette = "Set1") +
  theme_bw()
```

```{r}
## Proportional-odds (fixed effects only, no grower RE) via MASS::polr
## Expected to show near-zero variety effect (proportional-odds assumption fails).
## Full GLMM with grower random effect requires ordinal::clmm or glmmTMB >= 1.2.
if (requireNamespace("MASS", quietly = TRUE)) {
  fit_polr2 <- MASS::polr(
    rating ~ variety,
    weights = y,
    data    = DataSet14.2,
    Hess    = TRUE,
    method  = "logistic"
  )
  summary(fit_polr2)
}
```

## Choosing Between Ordinal and Nominal Models

| Criterion | Ordinal | Nominal |
|---|---|---|
| Category ordering | Natural order exists | No natural order |
| Parsimony | Fewer parameters | More parameters |
| Proportional odds | Must check | Not required |
| Example | Pain scores, quality ratings | Species, region |

## Key Takeaways

- Use **ordinal models** when categories have a natural ordering; they
  are more parsimonious under the proportional-odds assumption.
- Use **nominal (multinomial logistic) models** when no ordering exists.
- The **proportional-odds assumption** can be tested with a likelihood
  ratio test comparing the proportional-odds model to a partial
  proportional-odds model.
- `emmeans` provides marginal probability estimates for both model types.

## References

Stroup, W. W., Ptukhina, M., and Garai, S. (2024). *Generalized Linear Mixed Models: Modern Concepts, Methods and Applications* (2nd ed.). CRC Press.

Agresti, A. (2010). *Analysis of Ordinal Categorical Data*, 2nd ed. Wiley.
